TacticToe: Learning the Game of Tactics

### 2018-01-20 tactictoe_jv.tex
data from the latest experiments available for manual proof mining

%
%Techniques combining machine learning with translation to automated
%reasoning have recently become an important component of formal proof 
%assistants. 
%Such ``hammer'' techniques complement traditional proof
%assistant automation as implemented by tactics and decision
%procedures. n this paper we present a unified proof assistant
%automation approach which attempts to automate the selection of
%appropriate tactics and tactic-sequences combined with an optimized
%small-scale hammering approach.
%
%We implement the technique as a tactic-level automation for HOL4:
%TacticToe. It implements a modified A*-algorithm directly in HOL4 that
%explores different tactic-level proof paths, guiding their selection
%by learning from a large number of previous tactic-level
%proofs. Unlike the existing hammer methods, TacticToe avoids
%translation to FOL, working directly on the HOL level. By combining tactic 
%prediction and premise selection, TacticToe is able to re-prove 39\% of 7902 
%HOL4 theorems in 
%%less than
%5 seconds whereas the best single HOL(y)Hammer strategy solves 32\% in the 
%same 
%amount of time.

@@

\todo{include latest results: 66.56 percent of 7123 theorems in 60 seconds and
do a similar experiment for hh with minimization}

@@

\section{Case study}\todo{
  TG + RK: describe the two examples.
  RK: After implementation and small experiment is done (in about 1 week) 
      Test tactictoe find new examples on new goals and manual exploration mode.
      Give feedback.
}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.

@@

@@

%\begin{remark}
%The additional difficulty produced by the necessary induction is such that the 
%theorem could not be proven by \holyhammer.
%\end{remark}

\newcommand{\wed}{\begin{math}\wedge\end{math}}

Proofs discovered by \tactictoe discovered during evaluation.

for theorem \texttt{SURJ\_INJ\_INV} In 
theory \texttt{pred\_set})

\begin{example}\label{ex:casestudy}()

%val SURJ_INJ_INV = store_thm(
%  "SURJ_INJ_INV",
%  ``SURJ f s t ==> ?g. INJ g t s /\ !y. y IN t ==> (f (g y) = y)``,
\vspace{5mm}
\small

Goal
\begin{lstlisting}[language=SMLSmall,frame=tb]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}

@@

\end{example}

>>>>

%
%Techniques combining machine learning with translation to automated
%reasoning have recently become an important component of formal proof 
%assistants. 
%Such ``hammer'' techniques complement traditional proof
%assistant automation as implemented by tactics and decision
%procedures. n this paper we present a unified proof assistant
%automation approach which attempts to automate the selection of
%appropriate tactics and tactic-sequences combined with an optimized
%small-scale hammering approach.
%
%We implement the technique as a tactic-level automation for HOL4:
%TacticToe. It implements a modified A*-algorithm directly in HOL4 that
%explores different tactic-level proof paths, guiding their selection
%by learning from a large number of previous tactic-level
%proofs. Unlike the existing hammer methods, TacticToe avoids
%translation to FOL, working directly on the HOL level. By combining tactic 
%prediction and premise selection, TacticToe is able to re-prove 39\% of 7902 
%HOL4 theorems in 
%%less than
%5 seconds whereas the best single HOL(y)Hammer strategy solves 32\% in the 
%same 
%amount of time.
\todo{include latest results: 66.56 percent of 7123 theorems in 60 seconds and
do a similar experiment for hh with minimization}

@@

@@

\section{Case study: comparison with original proofs}\todo{
  TG + RK: describe the two examples.
  RK: After implementation and small experiment is done (in about 1 week) 
      Test tactictoe find new examples on new goals and manual exploration mode.
      Give feedback.
}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.

@@

%val SURJ_INJ_INV = store_thm(
%  "SURJ_INJ_INV",
%  ``SURJ f s t ==> ?g. INJ g t s /\ !y. y IN t ==> (f (g y) = y)``,

@@

%Name: 0 IS_GCD_UNIQUE
%proof length: 12
%proof time: 0.080819
%Statistics
%  infstep : 2539
%  nodes   : 396
%  maxdepth: 7
%Time: 40.309547
%  inferstep time: 30.15097
%  node_find time: 0.58993
%  node_crea time: 9.142918
%    pred time: 2.425785
%    thmpred time: 0.000101
%    mc time: 4.417004
%    inst time: 0.0
%Proof found: STRIP_TAC THEN REWRITE_TAC [fetch "gcd" "is_gcd_def"] THEN REPEAT 
%Cases THENL [metisTools.METIS_TAC [], REWRITE_TAC [SUC_NOT, ALL_DIVIDES_0, 
%compute_divides] THEN metisTools.METIS_TAC [numTheory.NOT_SUC], 
%metisTools.METIS_TAC [numTheory.NOT_SUC, DIVIDES_ANTISYM], metisLib.METIS_TAC 
%[LESS_EQUAL_ANTISYM, (fetch "divides" "DIVIDES_LE"), prim_recTheory.LESS_0], 
%metisTools.METIS_TAC [], (RW_TAC numLib.arith_ss) [(fetch "divides" 
%"divides_def")], metisTools.METIS_TAC [DIVIDES_ANTISYM], metisLib.METIS_TAC 
%[LESS_EQUAL_ANTISYM, (fetch "divides" "DIVIDES_LE"), prim_recTheory.LESS_0]]
%Org tac number: 1
%Original proof time: 0.0
%Recording proof time: 0.008556

Goal: \texttt{IS\_GCD\_UNIQUE} in theory \texttt{pred\_set}).
\begin{lstlisting}[language=SMLSmall,frame=tb]
!a b c d. is_gcd a b c wedge is_gcd a b d ==> (c = d)
\end{lstlisting}

Original \holfour proof script (2.03 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
PROVE_TAC[IS_GCD,DIVIDES_ANTISYM]
\end{lstlisting}

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
STRIP_TAC THEN 
REWRITE_TAC [fetch "gcd" "is_gcd_def"] THEN 
REPEAT Cases THENL 
  [METIS_TAC [], 
   REWRITE_TAC [SUC_NOT, ALL_DIVIDES_0, compute_divides] THEN 
     METIS_TAC [NOT_SUC], 
   METIS_TAC [NOT_SUC, DIVIDES_ANTISYM], 
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0], 
   METIS_TAC [], 
   RW_TAC numLib.arith_ss [divides_def], 
   METIS_TAC [DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0]]
\end{lstlisting}

tacictoe is resourceful.

Goal: \texttt{SURJ\_INJ\_INV} in theory \texttt{pred\_set})
\begin{lstlisting}[language=SMLSmall,frame=tb]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}

@@
